---
title: "Lefty Lemurs"
format: html
editor: visual
---

Loading library

```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
```

**Here's what I did to clean the data:**

Importing dataset

```{r}
L_PreClean <- read.csv("~/Documents/GitHub/Lefty_Lemurs/LeftyLemurs_Original.csv")
```

```{r}
summary(L_PreClean)
```

rename long column names to simpler ones

```{r}
L_PreClean <- rename(L_PreClean, Troop = Troop.............A..B..C..D...., Time = Time..H.M.S., Note = Focal.sample.note......................................................................................................................................................please.include.as.much.detail.as.possible...If.a.note.is.about.a.grooming.bout..for.example..include.stop.time.in.the.same.cell.)
```

remove commas in the category column

```{r}
L_PreClean$Category <- str_replace(L_PreClean$Category, ",", "")
```

Why on earth did it make it into a value instead of a new data frame??? I guess it doesn't matter because it still works...

Sometimes I wrote behaviors in all caps, and sometimes I didn't. I'm just going to put all of them in all caps because it's easier.

```{r}
L_PreClean$Category = toupper(L_PreClean$Category)
```

I have some data where I didn't actually record the limb used. Why did I do that? I'm going to delete the rows where I forgot to list the limb.

Only keep rows in categories that don't contain RH, LH, RF, and/or LF

```{r}
L_HFdata <-
  L_PreClean %>%
  filter(str_detect(Category, "RH|RF|LH|LF"))
```

Squeaky clean

Now I want to create a separate data frame that only contains behaviors from my ethogram:

```{r}
# do not run 
# lefty_justEthogram <-
# L %>%  select(Category) %>%
# filter(str_detect(Category, "BP|QP|L|AG|SG|F|E|CL|HC|R|G|"))
```

Why didn't it work

*This did not work* Would it be possible to split the "Category" column into 2 columns? 1 that says "Task" and another that says "Limb"?

"Limb" would have LH/RH/LF/RH and "Task" would have the other stuff

I think if I duplicate the column and rename it and then eliminate the limbs from "Task" and keep only the limbs in "Limbs" it might work.

Then I can rename the tasks to "Eating/Grooming/Locomotion/etc.

Duplicating "Category" and calling the new one "Limb"

```{r}
# do not run: L_HFdata$Limb <- L_PreClean$Category
# do not run: L_PreClean2 <- L_PreClean %>% relocate (Limb, .before = Note)
```

*This did not work*

Replacing limb with nothing

```{r}
# do not run: lefty_split <- L_PreClean2$Category <- str_replace(L_PreClean$Limb, "BP|QP|L|AG|SG|F|E|CL|HC|R|G|", "")

# do not run: lefty_fullsplit <- L_PreClean2$Limb <- str_replace(L_PreClean$Category, "LF|LH|RF|RH|", "")
```

Categories kept limbs because they include an R (like rest) or L (like locomotion) I think limb section stayed the same? I think it kept all the columns that contained the limbs, which was all of them ofc. I might be running the wrong code.

Okay I'm going to use find and replace in Excel to fix it because R doesn't know how to do anything apparently and I literally want to cry rn

Wait wtf why does it say 30,000+ rows columns

## okay I've spent 4 hours trying to get it to work and tbh this is a waste of time so I'm just going to export the behavior data as a CSV and finish cleaning it in Google Sheets

Writing as CSV bc my attempts at making limb column completely failed after working on it for 2 days

```{r}
# write_csv(L_PreClean,"Lefty_Lemurs.csv")
```

------------------------------------------------------------------------

**Here's what I did in Google Sheets with the CSV file** (I really wanted to do it in R but I could not figure it out and wasted hours and hours)

-   Deleted rows that didn't contain behaviors from my ethogram
-   Copied "Category" into 2 columns
-   In old Category column, deleted limb data
-   In new "Limb" data, kept only limb data (not behavior)
-   Deleted rows that had question marks in data
-   Deleted data where category was "Other"
-   Deleted data that had rows with both hands
-   Merged "F" and "E" into one category "E"
-   Looked through notes to change category when it was just "G". Most got changed to L, some got changed to R, a few got deleted
-   Added new column for H/F to say whether limb was hand or foot
-   Added new column for side to say whether limb was left or right

New categories - E (eating/foraging) - R (resting) (lemur is still) - L (locomotion) (active motion like leaping, climbing, or jumping) - CL (cross legs) - HC (hand clasp) - BP (bipedal walking) - QP (quadrupedal walking) - LA (land) - SG (self grooming) - AG (allo-grooming)

***Upload clean data*** (it was a nightmare of cleaning this data in R and Excel and took many terrible hours, of which I do not wish to relive)

```{r}
L <- read.csv("~/Documents/GitHub/Lefty_Lemurs/LeftyLemurs_Clean.csv")
```

**Analysis**

\~Hex codes for colors (to use in graphs)\~ Red: #C84E00 Orange: #E89923 Yellow: #FFD960 Light Green: #A1B70D Turquoise: #339898 Teal: #1D6363 Medium Blue: #005587 Light Blue: #0577B1 Purple: #993399

```{r}
head(L)
```

Let's look at how many total instances of each limb use there was

```{r}
table(L$Limb)
```

Omg

Left Hand: 225 Right Hand: 318 Left Foot: 32 Right Foot: 47

It looks like lemurs used their right hands a lot more than they used their left hands...

```{r}
 ggplot (L, aes(x= reorder (Limb, Limb, table ))) +
  geom_bar(color = "black", fill = "#012169") +
  theme_classic() +
  xlab("Limb") +
  ylab("Total Times Using Each Limb") +
  ggtitle("Total Limb Use Across Lemurs") +
  coord_flip()
```

Yep, there definitely appears to be a hand/foot preference

```{r}
ggplot (L, aes(x= reorder (Limb, Limb, table ))) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Limb") +
  ylab("Total Times Using Each Limb") +
  ggtitle("Total Limb Use Across Lemur Species") +
  coord_flip() +
  facet_wrap (~ Species, scales = "free")
```

```{r}
ggplot (L, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Limb Use Between Lemur Species") +
  facet_wrap (~ Species, scales = "free")
```

```{r}
ggplot (L, aes(x= reorder (Side, Side, table ))) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right") +
  ylab("Total Times Using Each Limb") +
  ggtitle("Limb Use Between Lemur Species") +
  coord_flip() +
  facet_wrap (~ Species, scales = "free")
```

Same thing but on the same scale (I got more data from sifakas as u can see)

```{r}
ggplot (L, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Limb Use Between Lemur Species") +
  facet_wrap (~ Species,)
```

👀 Sifakas have higher left hand use than the other 2 species. Sifakas overall seem more ambidextrous. However, this could be due to them foraging more (maybe all lemurs forage more with their left hands or something idk). Therefore I should separate it out by eating behaviors.

What about by sex

```{r}
ggplot (L, aes(x = reorder (Limb, Limb, table ))) +
  geom_bar(color = "black", fill = "#1D6363") +
  theme_classic() +
  xlab("Limb") +
  ylab("Total Times Using Each Limb") +
  ggtitle("Total Limb Use Between Sexes: Lemurs Are Not Lefties") +
  coord_flip() +
  facet_wrap (~ Sex, scales = "free")
```

This is the same thing but just divided by left and right side (left hand and left foot, etc. are in the same category)

```{r}
ggplot (L, aes(x= reorder (Side, Side, table ))) +
  geom_bar(color = "black", fill = "#1D6363") +
  theme_classic() +
  xlab("Left or Right") +
  ylab("Total Times Using Each Limb") +
  ggtitle("Total Limb Use Between Sexes: Lemurs Are Not Lefties") +
  coord_flip() +
  facet_wrap (~ Sex, scales = "free")
```

Definitely more right hand than left HAND use for both males and females. Not sure about foot use, but I didn't get much foot use data to begin with. Is there a way to exclude foot use?

Males might me a little less lateralized than females? Left hand use is also (slightly) higher in males like I thought it would be!

------------------------------------------------------------------------

```{r}
ggplot (L, aes(x= reorder (Category, Category, table ))) +
  geom_bar(color = "black", fill = "#0577B1") +
  theme_classic() +
  xlab("Type of Behavior") +
  ylab("Total Times Using Each Limb") +
  ggtitle("Behavior Across Sexes") +
  coord_flip() +
  facet_wrap (~ Sex,scales = "free")
```

Everyone eats a lot

Some calculations

```{r}
summary(L)

library(psych)

describeBy(L[c(4,1:12)], group = "Category")
# what does the 4 mean??
```

Making a DF with just hand grasp data

```{r}

# take out feet use data

l_hands <-
  L %>%
  select(H.F) %>%
  filter(str_detect(H.F, "H"))

# add the other stuff back
l_hands <- inner_join(l_hands, L, by = "H.F")

l_hands <- unique(l_hands)

# actually I also don't want to look at all the categories bc some were added just for fun and are probably not important

l_handstuff <-
  l_hands %>%
  select(Category) %>%
  filter(str_detect(Category, "E|L|R|SG"))


# WHY WON'T IT TAKE OUT LA

# add the other stuff back why did it even go away

l_handstuff <- inner_join(l_handstuff, l_hands, by = "Category")

# I can remove the H.F row because all of it is hand 

l_handstuff <- l_handstuff %>% select(-H.F)

# removing Limb row bc side says the same thing

l_handstuff <- l_handstuff %>% select(-Limb)

# removing mystery duplicates

l_handstuff <- unique(l_handstuff)


l_noLA <- 
l_handstuff %>% 
  filter(Category == "E"  |  Category == "L" |  Category == "R" |  Category == "SG" | Category != "LA" )

#wow I forgot you can do that

```

Compare hand use by species

```{r}
ggplot (l_noLA, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Hand Use Between Lemur Species") +
  facet_wrap (~ Species, scales = "free")
```

```{r}
ggplot (l_noLA, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("Type of Behavior") +
  ggtitle("Hand Use Between Behaviors") +
  facet_wrap (~ Category, scales = "free")
```

So it looks like lemurs mostly use their right hand for everything except for grooming! Interesting

Now I'm going to filter out the grooming data so it's all grasping

```{r}
l_grasps <-
l_noLA %>% 
  filter(Category == "E"  |  Category == "L" |  Category == "R")
```

Compare grasping hand by species

```{r}
ggplot (l_grasps, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Hand Use for Grasping Between Lemur Species") +
  facet_wrap (~ Species, scales = "free")
```

What about by sex?

```{r}
ggplot (l_grasps, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Hand Use for Grasping Between Sexes") +
  facet_wrap (~ Sex, scales = "free")
```

It doesn't look like there's really a difference

What about just food grasps?

```{r}
l_foodGrasps <-
l_grasps %>% 
  filter(Category == "E")
```

Compare grasping hand by species

```{r}
ggplot (l_foodGrasps, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Hand Use for Food Grasping Between Lemur Species") +
  facet_wrap (~ Species, scales = "free")
```

It really looks like sifakas are more ambidextrous when it comes to food grasping!

What about by individual?

```{r}
ggplot (l_foodGrasps, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Hand Use for Food Grasping Between Individuals") +
  facet_wrap (~ Focal, scales = "free")
```

It kind of looks like Sophia and Thrax are lefties!!! And Licinius is close... The other lemurs look like righties!

Look at them all on the same scale

```{r}
ggplot (l_foodGrasps, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Hand Use for Food Grasping Between Individuals") +
  facet_wrap (~ Focal)
```

Okay there isn't much data for Sophia so it could just be "noise" for her... But Thrax has a lot of data

But is all of this actually statistically significant Don't know

```{r}
ggplot (l_noLA, aes(x= Category)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Activity") +
  ylab("Occurrences") +
  ggtitle("Hand Use for Activities by Species") +
  facet_wrap (~ Species)
```

This just does occurrences. How can I get the Y-axis to be % of L compared to R?

Idk

```{r}
ggplot (l_foodGrasps, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Activity") +
  ylab("Occurrences") +
  ggtitle("Hand Used for Food Grasping Between Species") +
  facet_wrap (~ Species)
```

```{r}
ggplot (l_foodGrasps, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Hand Use for Grasping Between Individuals") +
  facet_wrap (~ Focal, scales = "free" )
```

```{r}
ggplot (l_foodGrasps, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Side") +
  ylab("Occurrences") +
  ggtitle("Hand Used for Food Grasping Between Species") +
  facet_wrap (~ Species)
```

```{r}
ggplot (L, aes(x= Focal)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Focal") +
  ylab("Occurrences") +
  ggtitle("# of Times Using Each Limb") +
  facet_wrap (~ Side)
```

```{r}
ggplot (l_foodGrasps, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Side") +
  ylab("Occurrences") +
  ggtitle("Hand Used for Food Grasping Between Sex") +
  facet_wrap (~ Sex, scales = "free")
```

Trying some statistical tests

```{r}
L %>% 
  group_by(Side) %>% 
  summarise (n = n()) %>%
  mutate(proportion = n / sum(n))

l_handstuff %>% 
  group_by(Side) %>% 
  summarise (n = n()) %>%
  mutate(proportion = n / sum(n))

binom.test(218, 305, p = .75, alternative = "two.sided")
```

Trying to do chi-squared test and contigency coefficients (did not work)

```{r}
# tbl_handGrasps <- as.data.frame(l_noLA) %>% 
  # group_by(Side, Sex) %>% 
  # summarize(qty = sum(Freq)) %>% 
  # ungroup() %>% 
 #  spread(key = Sex, value = qty)

# mat_handGrasps <- as.matrix(tbl_handGrasps[-1])
# rownames(mat_handGrasps) <- levels(tbl_handGrasps$Side)

# chisq.test(mat_handGrasps)
```

It says it can't because it doesnt understand "Freq"

```{r}
ggplot (l_grasps, aes(x= Side)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Hand Use for Grasping Between Sexes") +
  facet_wrap (~ Sex, scales = "free")
```

```{r}
#I echo: false
ggplot (L, aes(x= Limb)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Limb") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Limb Use Between Individuals") +
  facet_wrap (~ Focal)
```

Not a lot of foot use. It might not be very useful to include these data

Just look at bipedal locomotion

```{r}
l_feet <- 
L %>% 
  filter(Category == "BP")

ggplot (l_feet, aes(x= Limb)) +
  geom_bar(color = "black", fill = "#C84E00") +
  theme_classic() +
  xlab("Limb") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Foot Use Between Individuals") +
  facet_wrap (~ Focal)
```

I don't really see any overall patterns here. It looks like Furia leads with her right foor more than left and Thrax leads a little more with his left foot than right.

Does age impact hand use?

```{r}
ggplot(L, aes(x = Category, y = Age, color = Side)) + 
  geom_jitter()  +
  theme_classic() +
  xlab("Type of Behavior") +
  ylab("Age of Lemur") +
  ggtitle("Does Age Affect Limb Use?") + 
  scale_color_manual(values = c("#1D6363", "#E89923"))
```

Yeah I don't think it does lol

Trying to make the graph I yearn for

```{r}
ggplot (l_noLA, aes(x= Category, fill = Side)) +
  geom_bar(color = "black") +
  theme_classic() +
  xlab("Type of Behavior") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Limb Use Between Lemur Species") +
  facet_wrap (~ Species, scales = "free") + 
  scale_fill_manual(values = c("#1D6363", "#E89923"))
```

It's still not exactly what I want. I want the Y-axis to be % of grasps being with left hand, not total times used. Splitting by color makes it really hard to read, but Idk how else to do it :P

```{r}
ggplot (l_noLA, aes(x= Side, fill = interaction(Category, Side))) +
  geom_bar(color = "black") +
  theme_classic() +
  xlab("Left or Right Side") +
  ylab("# of Times Using Each Limb") +
  ggtitle("Limb Use Between Lemur Species") +
  facet_wrap (~ Species, scales = "free")
```

Wow that it was too complicated. Why did I dare make this monstrosity?

Okay, the landing and foot use data is fun and all, but the hand grasps it what I actually need to look at. I'm going to make that into a .CSV and analyze the data in JMP

```{r}
# write_csv (l_grasps, "HandGrasps.csv")

summary(l_grasps)
```

## Testing significance:

Species (significant)

```{r}
#| echo: false
table(l_grasps$Side, l_grasps$Species)
chisq.test(table(l_grasps$Side, l_grasps$Species))
```

Sex (not significant)

```{r}
#| echo: false
table(l_grasps$Side, l_grasps$Sex)
chisq.test(table(l_grasps$Side, l_grasps$Sex))
```

Focal (significant)

```{r}
#| echo: false
table(l_grasps$Side, l_grasps$Focal)
chisq.test(table(l_grasps$Side, l_grasps$Focal))
```

Category (not significant)

```{r}
#I echo: false
table(l_grasps$Side, l_grasps$Category)
chisq.test(table(l_grasps$Side, l_grasps$Category))
```

Age (significant)

```{r}
#I echo: false
table(l_grasps$Side, l_grasps$Age)
chisq.test(table(l_grasps$Side, l_grasps$Age))
```
